---
title: "Life Expectancy Data Analysis"
author: "STAT 420, Summer 2023, Maxim Aldaranov, Alice Fu, Scott Girten"
date: '08/04/23'
output:
  html_document: 
    theme: readable
    toc: yes
  pdf_document: default
urlcolor: cyan
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```

```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(GGally)
library(patchwork)
library(readr)
library(tidyverse)
library(lmtest)
library(kableExtra)
```


```{r helper-functions, include=FALSE}
high_vif = function(model){
  vif_vals = faraway::vif(model)
  vif_vals[vif_vals > 5]
}

calc_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

influential_proportion = function(model){
  mean(cooks.distance(model) > 4 / length(resid(model)))
}

rmse = function(actual, predicted){
  sqrt(mean((actual - predicted)^2))
}

outlier_proportion = function(model){
  mean(rstandard(model)[abs(rstandard(model)) > 2])
}
```

```{r diagnostics-function, include=FALSE}

diagnostics = function(model, pcol = 'grey', lcol = 'dodgerblue', alpha = 0.01, plotit = TRUE, testit = TRUE){
  
  # Create individual diagnostic components
  
  # Fitted vs Residuals plot
  p1 = ggplot(model, aes(x = .fitted, y = .resid)) +
    geom_point(color = pcol) +
    geom_hline(yintercept = 0, color = lcol) +
    theme_minimal() +
    labs(x = 'Fitted',
         y = 'Residuals',
         title = 'Fitted vs. Residuals Plot')

  # Normal Q-Q Plot
  p2 = ggplot(model, aes(sample = .resid)) +
    geom_qq(color = pcol) +
    geom_qq_line(color = lcol) +
    theme_minimal() +
    labs(x = 'Theoretical Quartiles',
         y = 'Sample Quartiles',
         title = 'Normal Q-Q Plot')
  
  # Combine plots side by side
  p_final = p1 + p2


  # p-value and decision 
  p_val = shapiro.test(model$residuals)$p.value
  decision = if_else(p_val < alpha, 'Reject', 'Fail to Reject')

  
  # Logic for which diagnostics should be returned by function
  output = list('p_val' = p_val, 'decision' = decision, 'plots' = p_final)
  
  if(plotit == TRUE & testit == TRUE){
    return(output)
  }
  else if(plotit == TRUE & testit == FALSE){
    return(output[3])
  }
  else if(plotit == FALSE & testit == TRUE){
    return(output[1:2])
  }
  
}
```


# Introduction

The purpose of this analysis is to understand which factors are prominent in affecting global life expectancy.  In an attempt to quantify these affects, we will fit a model that predicts the average life expectancy based on a variety of socioeconomic and healthcare-related factors. The appropriate data was found to be analyzed. Different statistical methods will be applied to the data in order to reach the best model for the given purpose. It will allow us to determine interactions between the average life expectancy and the various factors affecting it, hence, helping us to determine the variables that accurately predict the average life expectancy.

## Dataset Information

The data represents the research on the average life expectancy in 178 countries over a 15-year period (2000-2015) from the World Health Organization. It includes economic, social and healthcare related factors affecting the average life expectancy. The data contains `21` variables and `2,864` observations. It was retrieved from [Kaggle](https://www.kaggle.com/datasets/lashagoch/life-expectancy-who-updated).

- `Country` - country of the sample
- `Region` - region of the sample
- `Year` - year of the sample
- `infant_deaths` - the number of infant deaths per 1000 people
- `Under_five_deaths` - the number of deaths of people under 5 years old per 1000 people
- `Adult_mortality` - the number of deaths per 1000 people between the ages of 15 years of age to 60 years of age
- `Alcohol_consumption` - alcohol consumption that is recorded in liters of pure alcohol per capita with 15+ years old
- `Hepatitis_B` - % of coverage of Hepatitis B (HepB3) immunization among 1-year-olds
- `Measles` - % of coverage of Measles containing vaccine first dose (MCV1) immunization among 1-year-olds
- `BMI` - measure of nutritional status in adults. It is defined as a person's weight in kilograms divided by the square of that person's height in meters (kg/m2)
- `Polio` - % of 1-year-olds immunized against polio
- `Diphtheria` - % of of 1-year-olds who were immunized against diphtheria
- `Incidents_HIV` - the number of deaths caused by AIDS of the last 4-year-olds who were born alive per 1000 people
- `GDP_per_capita` - gross domestic product per capita
- `Population_min` - total population in millions
- `Thinness_ten_nineteen_years` - prevalence of thinness among adolescents aged 10-19 years. BMI < -2 standard deviations below the median.
- `Thinness_five_nine_years` - prevalence of thinness among children aged 5-9 years. BMI < -2 standard deviations below the median.
- `Schooling` - average years that people aged 25+ spent in formal education
- `Economy_status_Developed` - developed country or not
- `Economy_status_Developing` - developing country or not
- `Life_expectancy` - average life expectancy of both genders in different years from 2010 to 2015

# Methods 

## Data Cleaning

- Read the data from the file

```{r read the data}
data_raw = read.csv("Life-Expectancy-Data-Updated.csv")
```

- Check any missing data

```{r check for na values}
ifelse(sum(is.na(data_raw)) == 0, "The data is complete", "The data is incomplete")
```

- One of the `Economy_status_Developed` and `Economy_status_Developing` variables is redundant. They can be combined in one variable

```{r compare economy_status variables}
head(subset(data_raw, select = c("Economy_status_Developed", "Economy_status_Developing")), 5)
```

- We decide to keep `Economy_status_Developed` variable where the value `1` indicates a developed country and `0` indicates a country that is still developing. For simplicity, we renamed the `Economy_status_Developed` variable to `Developed`

```{r}
data = subset(data_raw, select = -c(Economy_status_Developing))
names(data)[names(data) == 'Economy_status_Developed'] = 'Developed'
```

- Since we are interested in a model predicting **global** average life expectancy omitted the variables `Country` and `Region`

```{r}
data = subset(data, select = -c(Country, Region)) #Year
```

- We convert *character* variable `Developed` and the *integer* variable `Year` to *factor* type in order to use them as categorical variable in the further analysis.

```{r}
data$Developed = as.factor(data$Developed)
data$Year = as.factor(data$Year)
```

## Cleaned Data Overview

- View the correlation matrix graph and table of the cleaned data

```{r cor-matrix, message=FALSE}
# takes a long time to load, will need this for the Methods Section
# ggpairs(subset(data, select = -c(Year, Developed)))
```

```{r remove-1, eval=FALSE, message=FALSE, include=FALSE}
# just using this for closer looks at the pairs

# ggpairs(subset(data, select = c(Life_expectancy, Infant_deaths, Under_five_deaths, Adult_mortality)))

# ggpairs(subset(data, select = c(Life_expectancy, Hepatitis_B, Measles, Polio, Diphtheria)))

# ggpairs(subset(data, select = c(Life_expectancy, Incidents_HIV, Population_mln, Schooling, GDP_per_capita)))
```



```{r transformation}
pairs_data = data.frame(Life_expectancy = data$Life_expectancy, GDP_per_capita = data$GDP_per_capita, log_GDP= log(data$GDP_per_capita), log2_GDP = log(log(data$GDP_per_capita)))

transform_GDP = ggpairs(pairs_data) 


pairs_data = data.frame(subset(data, select = c(Life_expectancy, Hepatitis_B)), log_Hepatitis_B = log(data$Hepatitis_B), poly_Hepatits_B = poly(data$Hepatitis_B, 2))

transform_HepB = ggpairs(pairs_data) 


pairs_data = data.frame(subset(data, select = c(Life_expectancy, Diphtheria)), log_Diphtheria = log(data$Diphtheria), poly_Diphtheria = poly(data$Diphtheria, 2))

transform_Diph = ggpairs(pairs_data)


pairs_data = data.frame(subset(data, select = c(Life_expectancy, Schooling)), log_ = log(data$Schooling), poly_ = poly(data$Schooling, 2))

transform_Schooling = ggpairs(pairs_data) 


pairs_data = data.frame(subset(data, select = c(Life_expectancy, Incidents_HIV)), log_ = log(data$Incidents_HIV), poly_ = poly(data$Incidents_HIV, 2))

transform_hiv = ggpairs(pairs_data) 


pairs_data = data.frame(subset(data, select = c(Life_expectancy, Under_five_deaths)), log_ = log(data$Under_five_deaths), poly_ = poly(data$Under_five_deaths, 2))

transform_under5 = ggpairs(pairs_data)
```


```{r transform-plots, message=FALSE}
transform_GDP # log2 > log >> x, only minor improvement from log to log2

# transform_HepB  # no improvements

# transform_Diph # no improvements

# transform_Schooling # no improvements

# transform_hiv # slight improvement with log?

transform_under5 # no improvements?

```


## Train and Test Data Separation

```{r ttsplit}
set.seed(42)
sample = sample(c(TRUE, FALSE), nrow(data), replace=TRUE, prob=c(0.7,0.3))
data_trn = data[sample, ]
data_tst = data[!sample, ]

```


## Additive Model

```{r cor1}
add_mod_full = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita), data = data_trn)
```

```{r collinearity reduction}
# initial full additive model
m1 = add_mod_full

# --- iteration 1 --- #

# check for high inflation factor variables
vif_vals = faraway::vif(m1)
vif_vals[vif_vals > 5] # Under_five_death has the highest vif

# check if Under_five_deaths is a useful predictor
m_under5 = lm(Under_five_deaths ~ . - GDP_per_capita + log(GDP_per_capita) - Life_expectancy, data = data_trn)
m_wo_under5 = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths, data = data_trn)
(cor1 = cor(resid(m_under5), resid(m_wo_under5))) # -0.2727 (low)

# remove Under_five_deaths
m1 = m_wo_under5


# --- iteration 2 --- #

# check for high inflation factor variables
vif_vals = faraway::vif(m1)
vif_vals[vif_vals > 5] # Diphtheria has the highest vif

# check if Diphtheria is a useful predictor
m_diph = lm(Diphtheria ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Life_expectancy, data = data_trn)
m_wo_dipth = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria, data = data_trn)
(cor1 = cor(resid(m_diph), resid(m_wo_dipth))) # -0.03076 (low)

# remove Diphtheria
m1 = m_wo_dipth


# --- iteration 3 --- #

# check for high inflation factor variables
vif_vals = faraway::vif(m1)
vif_vals[vif_vals > 5] # Thinness_ten_nineteen_years has the highest vif

# Check if Thinness_ten_nineteen_years is a useful predictor
m_thin1019 = lm(Thinness_ten_nineteen_years ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Life_expectancy, data = data_trn)
m_wo_thin1019 = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years, data = data_trn)
(cor1 = cor(resid(m_thin1019), resid(m_wo_thin1019))) # -0.02997 (low)

# remove Thinness_ten_nineteen_years
m1 = m_wo_thin1019

# --- iteration 4 --- #

# check for high inflation factor variables
vif_vals = faraway::vif(m1)
vif_vals[vif_vals > 5] # Infant_deaths has the highest vif

# Check if Infant_deaths is a useful predictor
m_infant = lm(Infant_deaths ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years - Life_expectancy, data = data_trn)
m_wo_infant = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years - Infant_deaths, data = data_trn)
(cor1 = cor(resid(m_infant), resid(m_wo_infant))) # -0.6504 (high)

# keep Infant_deaths
m1 = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years, data = data_trn)

# --- iteration 4 --- #

# check for high inflation factor variables
vif_vals = faraway::vif(m1)
vif_vals[vif_vals > 5] # Adult_mortality has the highest vif

# Check if Adult_mortality is a useful predictor
m_adult = lm(Adult_mortality ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years - Life_expectancy, data = data_trn)
m_wo_adult = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years - Adult_mortality, data = data_trn)
(cor1 = cor(resid(m_adult), resid(m_wo_adult))) # -0.826 (high)

# keep Adult_mortality
m1 = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years, data = data_trn)

# --- Final Model --- #
summary(m1)

add_mod_small = m1
```




## Backward AIC
```{r back-aic}
add_back_aic_mod = step(add_mod_small, direction = "backward", trace = FALSE)

# summary(add_back_aic_mod)
length(coef(add_back_aic_mod)) # 11
```


## Backward BIC

```{r back-bic}
n = length(resid(add_mod_small))

add_back_bic_mod = step(add_mod_small, direction = "backward", trace = FALSE, k = log(n))

# summary(add_back_bic_mod)
length(coef(add_back_bic_mod)) # 10

high_vif(add_back_bic_mod) # Under_five_deaths & Adult_mortality

bic_wo_adult = lm(Life_expectancy ~ Under_five_deaths + Alcohol_consumption + BMI + Incidents_HIV + log(GDP_per_capita) + Thinness_ten_nineteen_years + Schooling + Developed, data = data_trn)

adult_mort_mod = lm(Adult_mortality ~ Under_five_deaths + Alcohol_consumption + BMI + Incidents_HIV + log(GDP_per_capita) + Thinness_ten_nineteen_years + Schooling + Developed, data = data_trn)

cor(resid(bic_wo_adult), resid(adult_mort_mod)) # large

```


## Unusual Observations
```{r unusual-obs}
# rstandard
outlier_proportion(add_back_aic_mod) # 0.12102
outlier_proportion(add_back_bic_mod) # 0.11872

# Influence
influential_proportion(add_back_aic_mod) # 0.06297
influential_proportion(add_back_bic_mod) # 0.06697
```


## Liner Regression Assumptions

```{r}
diagnostics(add_back_bic_mod)
```


```{r bic-poly}
data_bic = data_trn %>% 
  select("Under_five_deaths", "Adult_mortality", "Alcohol_consumption", "BMI", "Incidents_HIV", "GDP_per_capita", "Thinness_ten_nineteen_years",
         "Schooling", "Developed", "Life_expectancy")
# pairs(data_bic, col = 'dodgerblue')

formula_bic = formula(add_back_bic_mod)

# 
add_back_bic_mod2 = lm(Life_expectancy ~ poly(Under_five_deaths, 2) + Adult_mortality + Alcohol_consumption + BMI +  
                         Incidents_HIV + log(GDP_per_capita) + Thinness_ten_nineteen_years + Schooling + Developed,
                       data = data_trn)

# diagnostics(add_back_bic_mod2)
# summary(add_back_bic_mod2)

anova(add_back_bic_mod, add_back_bic_mod2)
```

```{r formal-assumptions}

# BP test
bp_results = c(
  bptest(add_mod_small)$p.value[[1]],
  bptest(add_back_aic_mod)$p.value[[1]],
  bptest(add_back_bic_mod)$p.value[[1]], # e-31
  bptest(add_back_bic_mod2)$p.value[[1]] # e-45
)

# Shapiro test
shapiro_results = c(
  shapiro.test(add_mod_small$residuals)$p.value,
  shapiro.test(add_back_aic_mod$residuals)$p.value,
  shapiro.test(add_back_bic_mod$residuals)$p.value, # 0.0026
  shapiro.test(add_back_bic_mod2$residuals)$p.value # 0.01874
)

```




## Model Comparisons 
```{r comparisons}
# AIC
aics = c(
  AIC(add_mod_small), # 6930
  AIC(add_back_aic_mod), ## 6914
  AIC(add_back_bic_mod), # 6918
  AIC(add_back_bic_mod2)
)

# BIC
bics = c(
  BIC(add_mod_small), # 7098
  BIC(add_back_aic_mod), # 6981
  BIC(add_back_bic_mod), ## 6980
  BIC(add_back_bic_mod2)
)

# AdjR2 
adjR2s = c(
  summary(add_mod_small)$adj, ## 0.9793
  summary(add_back_aic_mod)$adj, ## 0.9793
  summary(add_back_bic_mod)$adj, # 0.9792
  summary(add_back_bic_mod2)$adj
)

# LOOCV RMSE
loocv_rmses = c(
  calc_loocv_rmse(add_mod_small), # 1.367
  calc_loocv_rmse(add_back_aic_mod), ## 1.362
  calc_loocv_rmse(add_back_bic_mod), # 1.364
  calc_loocv_rmse(add_back_bic_mod2)
)

```



```{r}
# Train/Test RMSE
train_error = c(
  rmse(data_trn$Life_expectancy, predict(add_mod_small, data_trn)),
  rmse(data_trn$Life_expectancy, predict(add_back_aic_mod, data_trn)),
  rmse(data_trn$Life_expectancy, predict(add_back_bic_mod, data_trn)),
  rmse(data_trn$Life_expectancy, predict(add_back_bic_mod2, data_trn))
)

test_error = c(
  rmse(data_tst$Life_expectancy, predict(add_mod_small, data_tst)),
  rmse(data_tst$Life_expectancy, predict(add_back_aic_mod, data_tst)),
  rmse(data_tst$Life_expectancy, predict(add_back_bic_mod, data_tst)),
  rmse(data_tst$Life_expectancy, predict(add_back_bic_mod2, data_tst))
)

```


# Results

```{r results-tables, echo=FALSE}

life_models = c("Additive Model", "Backward AIC Model", "Backward BIC Model", "Backward BIC Model V2")

bp_results = as.character(signif(bp_results), 2)
formal_tests = data.frame(life_models, bp_results, shapiro_results) 
colnames(formal_tests) = c("Model", "Breusch-Pagan Test P-value", "Shapiro-Wilk Test P-value")


life_results = data.frame(life_models, train_error, test_error)
colnames(life_results) = c("Model", "Train RMSE", "Test RMSE")


criterions = data.frame(life_models, aics, bics, adjR2s, loocv_rmses) 
colnames(criterions) = c("Model", "AIC", "BIC", "Adj R^2", "LOOCV RMSE")

knitr::kable(formal_tests)
knitr::kable(criterions)
knitr::kable(life_results)
```

# Discussion

# Appendix

```{r app-helper-functions, eval=FALSE}
high_vif = function(model){
  vif_vals = faraway::vif(model)
  vif_vals[vif_vals > 5]
}

calc_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

influential_proportion = function(model){
  mean(cooks.distance(model) > 4 / length(resid(model)))
}

rmse = function(actual, predicted){
  sqrt(mean((actual - predicted)^2))
}

outlier_proportion = function(model){
  mean(rstandard(model)[abs(rstandard(model)) > 2])
}
```

```{r app-diagnostics-function, eval=FALSE}

diagnostics = function(model, pcol = 'grey', lcol = 'dodgerblue', alpha = 0.01, plotit = TRUE, testit = TRUE){
  
  # Create individual diagnostic components
  
  # Fitted vs Residuals plot
  p1 = ggplot(model, aes(x = .fitted, y = .resid)) +
    geom_point(color = pcol) +
    geom_hline(yintercept = 0, color = lcol) +
    theme_minimal() +
    labs(x = 'Fitted',
         y = 'Residuals',
         title = 'Fitted vs. Residuals Plot')

  # Normal Q-Q Plot
  p2 = ggplot(model, aes(sample = .resid)) +
    geom_qq(color = pcol) +
    geom_qq_line(color = lcol) +
    theme_minimal() +
    labs(x = 'Theoretical Quartiles',
         y = 'Sample Quartiles',
         title = 'Normal Q-Q Plot')
  
  # Combine plots side by side
  p_final = p1 + p2


  # p-value and decision 
  p_val = shapiro.test(model$residuals)$p.value
  decision = if_else(p_val < alpha, 'Reject', 'Fail to Reject')

  
  # Logic for which diagnostics should be returned by function
  output = list('p_val' = p_val, 'decision' = decision, 'plots' = p_final)
  
  if(plotit == TRUE & testit == TRUE){
    return(output)
  }
  else if(plotit == TRUE & testit == FALSE){
    return(output[3])
  }
  else if(plotit == FALSE & testit == TRUE){
    return(output[1:2])
  }
  
}
```

Analysis written by Maxim Aldaranov, Alice Fu, and Scott Girten