---
title: "Life Expectancy Data Analysis"
author: "STAT 420, Summer 2023, Maxim Aldaranov, Alice Fu, Scott Girten"
date: '08/04/23'
output:
  html_document: 
    theme: readable
    toc: yes
    toc_float: true
    toc_depth: 4
  pdf_document: default
urlcolor: cyan
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2)
library(GGally)
library(patchwork)
library(readr)
library(tidyverse)
library(lmtest)
library(kableExtra)
```


```{r helper-functions, include=FALSE}
high_vif = function(model){
  vif_vals = faraway::vif(model)
  vif_vals[vif_vals > 5]
}

calc_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

influential_proportion = function(model){
  mean(cooks.distance(model) > 4 / length(resid(model)))
}

rmse = function(actual, predicted){
  sqrt(mean((actual - predicted)^2))
}

outlier_proportion = function(model){
  mean(rstandard(model)[abs(rstandard(model)) > 2])
}
```

```{r diagnostics-function, include=FALSE}

diagnostics = function(model, pcol = 'grey', lcol = 'dodgerblue', alpha = 0.01, plotit = TRUE, testit = TRUE){
  
  # Create individual diagnostic components
  
  # Fitted vs Residuals plot
  p1 = ggplot(model, aes(x = .fitted, y = .resid)) +
    geom_point(color = pcol) +
    geom_hline(yintercept = 0, color = lcol) +
    theme_minimal() +
    labs(x = 'Fitted',
         y = 'Residuals',
         title = 'Fitted vs. Residuals Plot')

  # Normal Q-Q Plot
  p2 = ggplot(model, aes(sample = .resid)) +
    geom_qq(color = pcol) +
    geom_qq_line(color = lcol) +
    theme_minimal() +
    labs(x = 'Theoretical Quartiles',
         y = 'Sample Quartiles',
         title = 'Normal Q-Q Plot')
  
  # Combine plots side by side
  p_final = p1 + p2


  # p-value and decision 
  p_val = shapiro.test(model$residuals)$p.value
  decision = if_else(p_val < alpha, 'Reject', 'Fail to Reject')

  
  # Logic for which diagnostics should be returned by function
  output = list('p_val' = p_val, 'decision' = decision, 'plots' = p_final)
  
  if(plotit == TRUE & testit == TRUE){
    return(output)
  }
  else if(plotit == TRUE & testit == FALSE){
    return(output[3])
  }
  else if(plotit == FALSE & testit == TRUE){
    return(output[1:2])
  }
  
}
```

```{r color palette, include=FALSE}
green_dark = '#66c2a5'
orange = '#fc8d62'
purple = '#8da0cb'
pink = '#e78ac3'
green_light = '#a6d854'

```



# Introduction

The purpose of this analysis is to understand which factors are prominent in affecting global life expectancy.  In an attempt to quantify these affects, we will fit a model that predicts the average life expectancy based on a variety of socioeconomic and healthcare-related factors. The appropriate data was found to be analyzed. Different statistical methods will be applied to the data in order to reach the best model for the given purpose. It will allow us to determine interactions between the average life expectancy and the various factors affecting it, hence, helping us to determine the variables that accurately predict the average life expectancy.

## Dataset Information

The data represents the research on the average life expectancy in 178 countries over a 15-year period (2000-2015) from the World Health Organization. It includes economic, social and healthcare related factors affecting the average life expectancy. The data contains `21` variables and `2,864` observations. It was retrieved from [Kaggle](https://www.kaggle.com/datasets/lashagoch/life-expectancy-who-updated).

- `Country` - country of the sample
- `Region` - region of the sample
- `Year` - year of the sample
- `infant_deaths` - the number of infant deaths per 1000 people
- `Under_five_deaths` - the number of deaths of people under 5 years old per 1000 people
- `Adult_mortality` - the number of deaths per 1000 people between the ages of 15 years of age to 60 years of age
- `Alcohol_consumption` - alcohol consumption that is recorded in liters of pure alcohol per capita with 15+ years old
- `Hepatitis_B` - % of coverage of Hepatitis B (HepB3) immunization among 1-year-olds
- `Measles` - % of coverage of Measles containing vaccine first dose (MCV1) immunization among 1-year-olds
- `BMI` - measure of nutritional status in adults. It is defined as a person's weight in kilograms divided by the square of that person's height in meters (kg/m2)
- `Polio` - % of 1-year-olds immunized against polio
- `Diphtheria` - % of of 1-year-olds who were immunized against diphtheria
- `Incidents_HIV` - the number of deaths caused by AIDS of the last 4-year-olds who were born alive per 1000 people
- `GDP_per_capita` - gross domestic product per capita
- `Population_min` - total population in millions
- `Thinness_ten_nineteen_years` - prevalence of thinness among adolescents aged 10-19 years. BMI < -2 standard deviations below the median.
- `Thinness_five_nine_years` - prevalence of thinness among children aged 5-9 years. BMI < -2 standard deviations below the median.
- `Schooling` - average years that people aged 25+ spent in formal education
- `Economy_status_Developed` - developed country or not
- `Economy_status_Developing` - developing country or not
- `Life_expectancy` - average life expectancy of both genders in different years from 2010 to 2015

# Methods 

## Data Cleaning

Before we begin the analysis, we performed basic data cleaning on the modeling data set.  After the data is read in we checked for `NULL` values in the data and looked at the top 10 rows of the data set to have a better understanding of the data in each column.   


```{r read data and check nulls}
data_raw = read.csv("Life-Expectancy-Data-Updated.csv")
ifelse(sum(is.na(data_raw)) == 0, "The data is complete", "The data is incomplete")
```

```{r data head}

slice_head(data_raw, n = 10) %>% 
  kbl() %>% 
  kable_styling() %>% 
  scroll_box(width = "100%")
```
<br>

The variables `Economy_status_Developed` and `Economy_status_Developing` variables are redundant - the columns represent the same binary data coded in opposite fashion.  We decided to keep `Economy_status_Developed` and for convenience will rename this column `Developed` for use in our analysis.  

```{r}
data = subset(data_raw, select = -c(Economy_status_Developing))
names(data)[names(data) == 'Economy_status_Developed'] = 'Developed'
```

Our primary interest in this analysis is to predict global life expectancy and Country or Regional differences would be beyond the scope of this analysis.  Since we are limiting the analysis to global life expectancy, we will remove the variables `Country` and `Region` from the data set for the rest of the analysis.   

```{r}
data = subset(data, select = -c(Country, Region)) #Year
```

After reviewing the data, we think it is appropriate to consider `Developed` and `Year` as factors and we have converted these columns from numeric to factor data types.  

```{r}
data$Developed = as.factor(data$Developed)
data$Year = as.factor(data$Year)
```


## Transformations

```{r cor-matrix, message=FALSE, fig.width=15, fig.height=15}
theme_set(theme_minimal())

p_corr = ggpairs(subset(data, select = -c(Year, Developed)), aes(color = orange, fill = orange, alpha = 0.5))
p_corr
```


We continued preparing our data set by looking for variables having a non-linear relationship with our response variable `Life_expectancy`.  We created a correlation matrix of all predictors and the response to visually understand the relationships in our data. Due to the large size of the correlation plot, we have decided to omit the correlation matrix in this report. However one variable in particular `GDP_per_capita` stood out as having a non-linear relationship with `Life-expectancy`.  We feel `GDP_per_capita` could benefit from a log transformation in our modeling process.  Below is a plot of `Life_expectancy` vs. `GDP_per_capita` highlighting this relationship. 


```{r life vs gdp}
p = ggplot(data, aes(x = GDP_per_capita, y = Life_expectancy, color = orange, fill = orange, alpha = 0.2)) + 
  geom_point()+
  theme_minimal() +
  theme(legend.position = 'none') + 
  labs(x = 'GDP per Capita',
       y = 'Life Expectancy',
       title = 'Life Expectancy vs. GDP per Capita')
p
```




```{r remove-1, eval=FALSE, message=FALSE, include=FALSE}
# just using this for closer looks at the pairs

# ggpairs(subset(data, select = c(Life_expectancy, Infant_deaths, Under_five_deaths, Adult_mortality)))

# ggpairs(subset(data, select = c(Life_expectancy, Hepatitis_B, Measles, Polio, Diphtheria)))

# ggpairs(subset(data, select = c(Life_expectancy, Incidents_HIV, Population_mln, Schooling, GDP_per_capita)))
```


```{r transformation, include=FALSE}
pairs_data = data.frame(Life_expectancy = data$Life_expectancy, GDP_per_capita = data$GDP_per_capita, log_GDP= log(data$GDP_per_capita), log2_GDP = log(log(data$GDP_per_capita)))

transform_GDP = ggpairs(pairs_data) 


pairs_data = data.frame(subset(data, select = c(Life_expectancy, Hepatitis_B)), log_Hepatitis_B = log(data$Hepatitis_B), poly_Hepatits_B = poly(data$Hepatitis_B, 2))

transform_HepB = ggpairs(pairs_data) 


pairs_data = data.frame(subset(data, select = c(Life_expectancy, Diphtheria)), log_Diphtheria = log(data$Diphtheria), poly_Diphtheria = poly(data$Diphtheria, 2))

transform_Diph = ggpairs(pairs_data)


pairs_data = data.frame(subset(data, select = c(Life_expectancy, Schooling)), log_ = log(data$Schooling), poly_ = poly(data$Schooling, 2))

transform_Schooling = ggpairs(pairs_data) 


pairs_data = data.frame(subset(data, select = c(Life_expectancy, Incidents_HIV)), log_ = log(data$Incidents_HIV), poly_ = poly(data$Incidents_HIV, 2))

transform_hiv = ggpairs(pairs_data) 


pairs_data = data.frame(subset(data, select = c(Life_expectancy, Under_five_deaths)), log_ = log(data$Under_five_deaths), poly_ = poly(data$Under_five_deaths, 2))

transform_under5 = ggpairs(pairs_data)
```


```{r transform-plots, message=FALSE, include=FALSE}
transform_GDP # log2 > log >> x, only minor improvement from log to log2

# transform_HepB  # no improvements

# transform_Diph # no improvements

# transform_Schooling # no improvements

# transform_hiv # slight improvement with log?

transform_under5 # no improvements?

```


## Train and Test Data Separation

Splitting our data set into training (70%) and testing (30%) data sets for use in the analysis.   

```{r ttsplit}
set.seed(42)
sample = sample(c(TRUE, FALSE), nrow(data), replace=TRUE, prob=c(0.7,0.3))
data_trn = data[sample, ]
data_tst = data[!sample, ]

```


## Modeling

### Additive Model

Our first model is an additive model of all of the predictor variables in our data set.  After fitting the model we looked at variables with a VIF greater than 5 to identify any variables that may be co-linear.  

```{r add_mod}
add_mod_full = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita), data = data_trn)
add_len = length(coef(add_mod_full))

vif = faraway::vif(add_mod_full)
vif = vif[vif > 5]
vif_table = tibble(Predictor = names(vif),
                   VIF = vif)

vif_table %>% 
  kbl(caption = 'Predictors with VIF > 5') %>% 
  kable_styling()
```
<br>


#### Variance Inflation Factor

For the variables with a high VIF (> 5), we calculated the partial correlation coefficient to determine if the predictor variable is correlated to the response variable and if so then remove the variable from our analysis.  Ultimately this process will remove the following variables:  
- `Under_five_deaths`  
- `Diphtheria`  
- `Thinness_ten_nineteen_years`  
- ``


```{r collinearity reduction}
# initial full additive model
m1 = add_mod_full

# --- iteration 1 --- #

# check for high inflation factor variables
vif_vals = faraway::vif(m1)
vif_vals = vif_vals[vif_vals > 5] # Under_five_death has the highest vif

# check if Under_five_deaths is a useful predictor
m_under5 = lm(Under_five_deaths ~ . - GDP_per_capita + log(GDP_per_capita) - Life_expectancy, data = data_trn)
m_wo_under5 = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths, data = data_trn)
cor = (cor1 = cor(resid(m_under5), resid(m_wo_under5))) # -0.2727 (low)

# remove Under_five_deaths
m1 = m_wo_under5


# --- iteration 2 --- #

# check for high inflation factor variables
vif_vals = faraway::vif(m1)
vif_vals = vif_vals[vif_vals > 5] # Diphtheria has the highest vif

# check if Diphtheria is a useful predictor
m_diph = lm(Diphtheria ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Life_expectancy, data = data_trn)
m_wo_dipth = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria, data = data_trn)
cor = (cor1 = cor(resid(m_diph), resid(m_wo_dipth))) # -0.03076 (low)

# remove Diphtheria
m1 = m_wo_dipth


# --- iteration 3 --- #

# check for high inflation factor variables
vif_vals = faraway::vif(m1)
vif_vals = vif_vals[vif_vals > 5] # Thinness_ten_nineteen_years has the highest vif

# Check if Thinness_ten_nineteen_years is a useful predictor
m_thin1019 = lm(Thinness_ten_nineteen_years ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Life_expectancy, data = data_trn)
m_wo_thin1019 = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years, data = data_trn)
cor = (cor1 = cor(resid(m_thin1019), resid(m_wo_thin1019))) # -0.02997 (low)

# remove Thinness_ten_nineteen_years
m1 = m_wo_thin1019

# --- iteration 4 --- #

# check for high inflation factor variables
vif_vals = faraway::vif(m1)
vif_vals = vif_vals[vif_vals > 5] # Infant_deaths has the highest vif

# Check if Infant_deaths is a useful predictor
m_infant = lm(Infant_deaths ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years - Life_expectancy, data = data_trn)
m_wo_infant = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years - Infant_deaths, data = data_trn)
cor = (cor1 = cor(resid(m_infant), resid(m_wo_infant))) # -0.6504 (high)

# keep Infant_deaths
m1 = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years, data = data_trn)

# --- iteration 5 --- #

# check for high inflation factor variables
vif_vals = faraway::vif(m1)
vif_vals = vif_vals[vif_vals > 5] # Adult_mortality has the highest vif

# Check if Adult_mortality is a useful predictor
m_adult = lm(Adult_mortality ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years - Life_expectancy, data = data_trn)
m_wo_adult = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years - Adult_mortality, data = data_trn)
cor = (cor1 = cor(resid(m_adult), resid(m_wo_adult))) # -0.826 (high)

# keep Adult_mortality
m1 = lm(Life_expectancy ~ . - GDP_per_capita + log(GDP_per_capita) - Under_five_deaths - Diphtheria - Thinness_ten_nineteen_years, data = data_trn)

# --- Final Model --- #
#summary(m1)

add_mod_small = m1
```




### Backward AIC

Our next step in our model building process was to take our initial additive model and perform step-wise backward selection using AIC. 

```{r back-aic}
add_back_aic_mod = step(add_mod_small, direction = "backward", trace = FALSE)

# Do we need to check VIF here too?

# summary(add_back_aic_mod)
aic_len = length(coef(add_back_aic_mod)) # 11
```


### Backward BIC

Finally, we fitted a model using the initial additive model and performed step-wise backward selection using BIC. Two variables `Infant_deaths` and `Adult_mortality` displayed a high VIF and were tested to see if a low partial correlation coefficient could allow us to eliminate the variable from the model.  The partial correlation coefficient was relatively large and the decision was made to keep the variables in the model.  

```{r back-bic}
n = length(resid(add_mod_small))

add_back_bic_mod = step(add_mod_small, direction = "backward", trace = FALSE, k = log(n))

bic_len = length(coef(add_back_bic_mod)) # 10

vif_val = high_vif(add_back_bic_mod) # Infant_deaths & Adult_mortality

bic_wo_adult = lm(Life_expectancy ~ Infant_deaths + Alcohol_consumption + BMI + Incidents_HIV + log(GDP_per_capita) +
                    Thinness_ten_nineteen_years + Schooling + Developed, data = data_trn)

adult_mort_mod = lm(Adult_mortality ~ Infant_deaths + Alcohol_consumption + BMI + Incidents_HIV + log(GDP_per_capita) +
                    Thinness_ten_nineteen_years + Schooling + Developed, data = data_trn)

cor = cor(resid(bic_wo_adult), resid(adult_mort_mod)) # large ~ 0.82

```


## Unusual Observations
```{r unusual-obs}
# rstandard
outlier_proportion(add_back_aic_mod) # 0.12102
outlier_proportion(add_back_bic_mod) # 0.11872

# Influence
influential_proportion(add_back_aic_mod) # 0.06297
influential_proportion(add_back_bic_mod) # 0.06697
```


## Plots and Tables for the Results Section

```{r diagnostic plots for BIC model}
p_diag_bic = diagnostics(add_back_bic_mod, pcol = green_light, lcol = pink, testit = FALSE)

```


```{r BP and Shapiro Table}
#BP and Shapiro test comparison
# BP test
bp_results = c(
  bptest(add_mod_small)$p.value[[1]],
  bptest(add_back_aic_mod)$p.value[[1]],
  bptest(add_back_bic_mod)$p.value[[1]] # e-31
)

# Shapiro test
shapiro_results = c(
  shapiro.test(add_mod_small$residuals)$p.value,
  shapiro.test(add_back_aic_mod$residuals)$p.value,
  shapiro.test(add_back_bic_mod$residuals)$p.value # 0.0026
)

bp_shap_table = tibble(Model = c('Additive Model', 'AIC Model', 'BIC Model'),
                       `BP Test p-value` = bp_results,
                       `Shapiro-Wilk Test p-value` = shapiro_results)

```


```{r Model performance comparison}
# Training comparison for all models
# AIC
aics = c(
  AIC(add_mod_small), # 6930
  AIC(add_back_aic_mod), ## 6914
  AIC(add_back_bic_mod) # 6918
)

# BIC
bics = c(
  BIC(add_mod_small), # 7098
  BIC(add_back_aic_mod), # 6981
  BIC(add_back_bic_mod) ## 6980
)

# AdjR2 
adjR2s = c(
  summary(add_mod_small)$adj, ## 0.9793
  summary(add_back_aic_mod)$adj, ## 0.9793
  summary(add_back_bic_mod)$adj # 0.9792
)

# LOOCV RMSE
loocv_rmses = c(
  calc_loocv_rmse(add_mod_small), # 1.367
  calc_loocv_rmse(add_back_aic_mod), ## 1.362
  calc_loocv_rmse(add_back_bic_mod) # 1.364
)

trn_comp = tibble(Model = c('Additive Model', 'AIC Model', 'BIC Model'),
                  `Number of Predictors` = c(add_len, aic_len, bic_len),
                  AIC = aics,
                  BIC = bics,
                  `Adjusted R^2` = adjR2s,
                  `LOOCV RMSE` = loocv_rmses)

```



```{r training and testing rmse}
# Training and testing comparison for all models
# Train/Test RMSE
train_error = c(
  rmse(data_trn$Life_expectancy, predict(add_mod_small, data_trn)),
  rmse(data_trn$Life_expectancy, predict(add_back_aic_mod, data_trn)),
  rmse(data_trn$Life_expectancy, predict(add_back_bic_mod, data_trn))
)

test_error = c(
  rmse(data_tst$Life_expectancy, predict(add_mod_small, data_tst)),
  rmse(data_tst$Life_expectancy, predict(add_back_aic_mod, data_tst)),
  rmse(data_tst$Life_expectancy, predict(add_back_bic_mod, data_tst))
)

trn_tst_comp = tibble(Model = c('Additive Model', 'AIC Model', 'BIC Model'),
                      `Training RMSE` = train_error,
                      `Testing RMSE` = test_error)

```


# Results

**Need to add text to results, but thought this might be good outline**

*Compare training results for the 3 models*

```{r, echo=FALSE}
trn_comp %>% 
  kbl(caption = 'Model Comparison from Training') %>% 
  kable_styling()
```

*Compare BP and Shapiro tests*
```{r echo=FALSE}
bp_shap_table %>% 
  kbl(caption = 'BP and Shapiro Test Comparison') %>% 
  kable_styling()
```


*Compare training and testing performance*
```{r echo=FALSE}
trn_tst_comp %>% 
  kbl(caption = 'Training and Testing Comparison') %>% 
  kable_styling()
```


*Diagnostic plots for BIC model*
```{r echo=FALSE}
p_diag_bic
```




# Discussion




# Appendix


### Libraries Used in the Analysis
```{r message=FALSE, warning=FALSE, eval=FALSE}
library(ggplot2)
library(GGally)
library(patchwork)
library(readr)
library(tidyverse)
library(lmtest)
library(kableExtra)
```


### Helper Functions
```{r app-helper-functions, eval=FALSE}
high_vif = function(model){
  vif_vals = faraway::vif(model)
  vif_vals[vif_vals > 5]
}

calc_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

influential_proportion = function(model){
  mean(cooks.distance(model) > 4 / length(resid(model)))
}

rmse = function(actual, predicted){
  sqrt(mean((actual - predicted)^2))
}

outlier_proportion = function(model){
  mean(rstandard(model)[abs(rstandard(model)) > 2])
}
```

### Diagnostic Functions
```{r app-diagnostics-function, eval=FALSE}

diagnostics = function(model, pcol = 'grey', lcol = 'dodgerblue', alpha = 0.01, plotit = TRUE, testit = TRUE){
  
  # Create individual diagnostic components
  
  # Fitted vs Residuals plot
  p1 = ggplot(model, aes(x = .fitted, y = .resid)) +
    geom_point(color = pcol) +
    geom_hline(yintercept = 0, color = lcol) +
    theme_minimal() +
    labs(x = 'Fitted',
         y = 'Residuals',
         title = 'Fitted vs. Residuals Plot')

  # Normal Q-Q Plot
  p2 = ggplot(model, aes(sample = .resid)) +
    geom_qq(color = pcol) +
    geom_qq_line(color = lcol) +
    theme_minimal() +
    labs(x = 'Theoretical Quartiles',
         y = 'Sample Quartiles',
         title = 'Normal Q-Q Plot')
  
  # Combine plots side by side
  p_final = p1 + p2


  # p-value and decision 
  p_val = shapiro.test(model$residuals)$p.value
  decision = if_else(p_val < alpha, 'Reject', 'Fail to Reject')

  
  # Logic for which diagnostics should be returned by function
  output = list('p_val' = p_val, 'decision' = decision, 'plots' = p_final)
  
  if(plotit == TRUE & testit == TRUE){
    return(output)
  }
  else if(plotit == TRUE & testit == FALSE){
    return(output[3])
  }
  else if(plotit == FALSE & testit == TRUE){
    return(output[1:2])
  }
  
}
```

Analysis written by Maxim Aldaranov, Alice Fu, and Scott Girten